### 2025-04-09 chapters/1-qcd.tex
T-1 day to initial submission, nearly complete draft up to citations and small comments.

\epigraph{
        When you ask what are electrons and protons I ought to answer that this question is not a profitable one to ask and does not really have a meaning.
    %
    The important thing about electrons and protons is not what they are but how they behave, how they move.
    %
    I can describe the situation by comparing it to the game of chess.
    %
    In chess, we have various chessmen [sic], kings, knights, pawns and so on.
    %
    If you ask what a chessman is, the answer would be that it is a piece of wood, or a piece of ivory, or perhaps just a sign written on paper, or anything whatever.
    %
    It does not matter.
    %
    Each chessman has a characteristic way of moving and this is all that matters about it.
    %
    The whole game of chess follows from this way of moving the various chessmen.
}{--Paul Adrien Maurice Dirac}

@@

\begin{figure}[]
    {
        \hspace{-1.7cm}
        \input{includes/timeline.tikz}
    }
    \caption{
        A rough timeline of the history of \gls{qcd}.
        %
        \sam{Not complete! Many particles, as well as QGP, missing! May want to change notation, e.g. for particle discoveries}
    }
    \label{fig:qcd-timeline}
\end{figure}

@@

\begin{itemize}
    \item
        you are human -- or perhaps a web-scraping artificial intelligence -- that deserves to know both where you are (the universe) and what you are made of (mostly quarks and gluons, by mass);

@@

    % Caption
    \caption{
        Energy-weighted correlators capture correlations between high-energy structures within jets, such as particles (black lines) or subjets (purple cones), as probed by the resolved 3-point energy correlators of \Chap{ewocs} visualized by the colored disks above the jet.
        %
        The higher energy regions of the cartoon jet depicted here are represented by higher values of the associated resolved energy correlator, and darker colors in the associated colored disks.
    }
    % Figure Label
    \label{fig:picturebook_eventshapes}
\end{sourcefigure}

@@

In \Chap{jets}, we will discuss a beautiful complication to the story:
%
the high-energy quarks and gluons which emerge when we break apart a proton themselves contain more quarks and gluons%
\footnote{
    Hence the name ``Particles Inside Particles'' for this thesis.
}.
%
The resulting fractal-like structures emerging from scattered quarks and gluons, visualized in \Fig{picturebook_jets}, are called \glspl{jet}.
%
In \Sec{partonshower}, we discuss and roughly derive the partonic cascade model of energy flow within the quark, and of jet formation.
%
Unfortunately, experiments are messy, and to understand the behavior of partons we need to see through a contaminating haze of low-energy pollution, discussed in \Sec{low-energy-pollution}.

>>>>

\epigraph{
        When you ask what are electrons and protons I ought to answer that this question is not a profitable one to ask and does not really have a meaning.
    %
    The important thing about electrons and protons is not what they are but how they behave, how they move.
    %
    I can describe the situation by comparing it to the game of chess.
    %
    In chess, we have various chessmen [sic], kings, knights, pawns and so on.
    %
    If you ask what a chessman is, the answer would be that it is a piece of wood, or a piece of ivory, or perhaps just a sign written on paper, or anything whatever.
    %
    It does not matter.
    %
    Each chessman has a characteristic way of moving and this is all that matters about it.
    %
    The whole game of chess follows from this way of moving the various chessmen.
}{---Paul Adrien Maurice Dirac}

@@

\begin{figure}[]
    {
        \hspace{-1.7cm}
        \input{includes/timeline.tikz}
    }
    \caption{
        A rough timeline of the history of \gls{qcd}.
        %
        \sam{Not complete! Many particles, as well as QGP, missing! May want to change notation, e.g. for particle discoveries}
        %
        \sam{Noah recommends lines connecting things to dates...}
    }
    \label{fig:qcd-timeline}
\end{figure}

@@

\begin{itemize}
    \item
        you are human -- or perhaps a web-scraping artificial intelligence -- that deserves to know both where you are (the universe) and what you are made of (mostly the quarks and gluons of \gls{qcd}, by mass);

@@

    % Caption
    \caption{
        Energy-weighted correlators capture correlations between high-energy structures within jets, such as particles (purple/black lines) or subjets (purple cones).
        %
        The explicit example of the resolved 3-point energy correlators of \Chap{ewocs}, compute using real collider data, are shown as colored disks above the jet to visualize energy-weighted correlations at different angular scales.
        %
        The higher energy regions of the cartoon jet depicted here are represented by higher values of the associated resolved energy correlator, and darker colors in the associated colored disks.
    }
    % Figure Label
    \label{fig:picturebook_eventshapes}
\end{sourcefigure}

@@

In \Chap{jets}, we will discuss a beautiful complication to the story:
%
the high-energy quarks and gluons which emerge when we break apart a proton themselves contain more quarks and gluons%
\footnote{
    Hence the name ``Particles Inside Particles'' for this thesis.
}.
%
The resulting fractal-like structures emerging from scattered quarks and gluons, visualized in \Fig{picturebook_jets}, are called \glspl{jet}.
%
In \Sec{parton-shower}, we discuss and roughly derive the partonic cascade model of energy flow within the quark, and of jet formation.
%
Unfortunately, experiments are messy, and to understand the behavior of partons we need to see through a contaminating haze of low-energy pollution, discussed in \Sec{low-energy-pollution}.
